\chapter*{Acknowledgements}

One of my primary drivers for this work has been to contribute something that would be useful to other people.
If I am even a bit successful in this, it is thanks to the collaboration, discussions and support from my
advisors, colleagues, friends, and family.
This section is my attempt in thanking each and every one of them.

\small
It feels right to start by thanking my family. I want to thank my father, Kostas, for passing on his life values to me from an early age,
and by that making me who I am today.
To my mother, Georgia, and my sister, Christina, thank you for your care and support at my every step.

\bigskip
This work would not have been possible without the guidance, help and support of my advisors, Marc Shapiro and Brad King.
I am thankful to them for dedicating much of their time for this work,
and for always being available and helpful when I needed help.
Marc, thank you for inspiring me and improving this work by always pointing out some previously unidentified
design flaw or incorrect assumption at my every idea.
Brad, thank you for teaching me something new about systems, their use, research, and even fluid dynamics at our every
discussion.
Thank you to both for teaching me how to do systems research and how to communicate this research so that it can useful
to others.

\bigskip
I would like to thank the reviewers, S{\'e}bastien Monnet and Nuno Pregui\c{c}a for dedicating the time
to evaluate this thesis.
I also thank the examiners, Bernd Amann, Bettina Kemme, Themis Palpanas and Masoud Saeida Ardekani for accepting to be
part of my thesis committee.
A special thanks to S{\'e}bastien and Themis for also accepting to be part of the thesis monitoring committee,
and for offering their valuable feedback.

\bigskip
Research is not a solitary endeavour.
Good research is only made possible through discussion and collaboration.
I am fortunate to have been working among two great teams: Scality and the LIP6 laboratory.

I want to thank my colleagues at the Delys team.
Alejandro, Beno\^it, Francis, Ilyas, Jonathan, Laurent, Michal, Paolo, Saalik, Sara, Sreeja, Vincent, Vinh, thank you
for all the interesting discussions we have had over the years, and the time spent together.

Special thanks to Sara Hamouda for all her inputs, help and support with this work, and for being a great co-author.

// too much?
Some special nods to
Alejandro for the lunches at the Chinese restaurant,
Ilyas for the Wednesday night runs,
Jonathan whose the never-ending ``I have something interesting to show you'' have been the source of interesting and fun conversations,
Saalik for the interesting discussions on technology, football and pop culture over lunch and coffee,
and Sreeja for being someone I look up to as a researcher and a human being.
//

I would also like to thank the other Ph.D. students and interns at the LIP6 laboratory
Alexandre, Antoine, Arnaud, C\'edric, Damien, Daniel, Darius, Florent, Guillaume, Hakan, Jo\~ao Paulo, Lucas, Ludovic, Marjorie, Maxime, Redha and S\'ebastien,
as well as the senior researchers, Jonathan, Julien, Luciana and Pierre, with whom I have shared my days at the laboratory.

I am grateful to the Scality family. I want to thank the Ironman team for welcoming me early on, and Vinh Tao and
Vianney Rancurel for guiding me in my first steps in this thesis.
I would like to thank the File Squad for welcoming me more recently, and for supporting and teaching me about working within
an engineering team, and building a great product.
A special thanks to Aline and Marina for supporting me through the challenges of moving to France.
Thank you to Charles, Dogara, Duhamel, Florent, Greg, Imane, Jordi, Romain, Roxanne and St\'ephanie for the workout
sessions.
Also thanks to Alexandre, Anurag, Jordi and Partick for introducing me to bouldering.

In addition, I would like to thank all the members of the LightKone European project for the exchanges and collaboration.

\bigskip
Georgios Goumas, Vangelis Koukis and Nectarios Koziris from the National Technical University of Athens in Greece inspired
me and sparked my interest for understanding and building computing systems.
I also want to thank Stefanos Gerangelos for introducing me to the path of research, and guiding me through the first steps.

\bigskip
I am grateful to my close friends for accompanying me through this journey, and for supporting me,
each in their own way,
Aggeliki, Alexandros, Andreas, Argyro, Faidra, Giorgos K., Giorgos S., Giorgos V., Iliana, Kostis, Margarita, Maria and Thanos.

Finally, I owe a great deal to Kalliopi for her patience and understanding, for always being there to listen and for
her ability to lift my spirits no matter the circumstances.
