\chapter*{Acknowledgements}

One of my primary drivers for this work has been to contribute something that would be useful to other people.
If I am even a bit successful in this, it is thanks to the collaboration, discussions and support from my
advisors, colleagues, friends, and family.
This section is my attempt in thanking each and every one of them.

\small
It feels right to start by thanking my family. To my father, Kostas, thank you for passing on your life values to me,
and by that making me who I am today. I hope I am living up to your expectations.
To my mother, Georgia, and my sister, Christina, thank you for your care and support at my every step.

\bigskip
This work would not have been possible without the guidance, help and support of my advisors, Marc Shapiro and Brad King.
I am thankful to them for always being patient, helpful, and insightful.
Marc, thank you for all the time you dedicated discussing this work,
refining the ideas, and challenging me to improve.
Brad, thank you for teaching me something new about systems, their use, research, and even fluid dynamics in our every
discussion.
Thank you to both for teaching me how to do systems research and how to communicate this research so that it can useful
to others.

\bigskip
I would like to thank the reviewers, S{\'e}bastien Monnet and Nuno Pregui\c{c}a for dedicating the time
to evaluate this thesis.
I also thank the examiners, Bernd Amann, Bettina Kemme, Themis Palpanas and Masoud Saeida Ardekani for accepting to be
part of my thesis committee.
A special thanks to S{\'e}bastien and Themis for also accepting to be part of the thesis monitoring committee,
and for offering their valuable feedback.

\bigskip
Research is not a solitary endeavour.
Good research is only made possible through discussion and collaboration.
I am fortunate to have been working among two great teams: Scality and the LIP6 laboratory.

I want to thank my colleagues at the Delys team.
Alejandro Tomsic, Beno\^it Martin, Francis Laniel, Ilyas Toumlilt, Jonathan Sid-Otmane, Laurent Prosperi, Michael Rudek, Paolo Viotti, Saalik Hatia, Sara Hamouda, Sreeja Nair, Vincent Vallade and Vinh Tao Thanh, thank you
for all the interesting discussions we have had over the years, and the time spent together.

Special thanks to Sara Hamouda for all her inputs, help and support with this work, and for being a great co-author.

Some special nods to
Alejandro for the lunches at the Chinese restaurant,
Ilyas for the Wednesday night runs,
Jonathan whose the never-ending ``I have something interesting to show you'' were the beginning of fascinating endeavours,
Saalik for his inexhaustible knowledge on technology, pop culture and more that always led to interesting discussions,
and Sreeja for being someone I look up to as a researcher and a human being.

I would also like to thank the other Ph.D. students and interns at the LIP6 laboratory
Alexandre, Antoine, Arnaud, C\'edric, Damien, Daniel, Darius, Florent, Guillaume, Hakan, Jo\~ao Paulo, Lucas, Ludovic, Marjorie, Maxime, Redha and S\'ebastien,
as well as the senior researchers, Jonathan, Julien, Luciana and Pierre, with whom I have shared my days at the laboratory.

I am grateful to the Scality family. I want to thank the Ironman team for welcoming me early on, and Vinh Tao and
Vianney Rancurel for guiding me in my first steps in this thesis.
I would like to thank the File Squad for welcoming me more recently, and for supporting and teaching me about working within
an engineering team, and building a great product.
A special thanks to Aline and Marina for supporting me through the challenges of moving to France.
Thank you to Charles, Dogara, Duhamel, Florent, Greg, Imane, Jordi, Romain, Roxanne and St\'ephanie for the workout
sessions.
Thank you to Alexandre, Anurag, Jordi and Partick for introducing me to bouldering.
Also special thanks to Lam for all the discussions and the time we shared.

In addition, I would like to thank all the members of the LightKone European project for the exchanges and collaboration.

\bigskip
Georgios Goumas, Vangelis Koukis and Nectarios Koziris from the National Technical University of Athens in Greece inspired
me and sparked my interest for understanding and building computing systems.
I also want to thank Stefanos Gerangelos for introducing me to the path of research, and guiding me through the first steps.

\bigskip
I am grateful to my close friends for accompanying me through this journey, and for supporting me,
each in their own way,
Aggeliki, Alexandros, Andreas, Argyro, Faidra, Giorgos K., Giorgos S., Giorgos V., Iliana, Kostis, Margarita, Maria and Thanos.

Finally, I owe a great deal to Kalliopi for her patience and understanding, for always being there to listen and for
her ability to lift my spirits no matter the circumstances.
